<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Figures for the MBR Manuscript}
-->

# Figures for the MBR Manuscript
This vignette produces the graphs included in the initial MBR manuscript.

```{r set_options, echo=FALSE, results='hide'}
require(knitr)
opts_chunk$set(
    comment=NA, 
    tidy=FALSE,
    fig.width=6.5, 
    fig.height=1.6,
    fig.path='figure_rmd/',
    dpi=100
)

# options(markdown.HTML.header = system.file("misc", "vignette.css", package = "knitr"))
# options(markdown.HTML.header = system.file("misc", "vignette.css", package = "REDCapR"))
# options(markdown.HTML.header = file.path(devtools::inst("REDCapR"), "misc", "vignette.css"))

options(width=120) #So the output is 50% wider than the default.
```

## Figure 2: Cartesian Rolling
Smoothed monthly birth rates (General Fertility Rates; GFRs) for Oklahoma County, 1990-1999, plotted in a linear plot.  The top plot shows the connected raw data with a February smoother; the middle plot shows smoothing with a 12-month moving average, blue/green line, superimposed on a February smoother, red line); the bottom plot shows the smoothers and confidence bands, which are H-spreads defined using the distribution of GFR's for the given month and 11 previous months.

First, some R packages are loaded, and some variables and functions are defined.
```{r Definitions}
library(Wats)
library(grid)
library(plyr) 
library(ggplot2) 
library(boot) 

changeMonth <- as.Date("1996-02-15")

vpLayout <- function(x, y) { viewport(layout.pos.row=x, layout.pos.col=y) }

fullSpread <- function( scores) { 
  return( range(scores) ) #A new function isn't necessary.  It's defined just to be consistent.
}
hSpread <- function( scores) { 
  return( quantile(x=scores, probs=c(.25, .75)) ) 
}
seSpread <- function( scores) { 
  return( mean(scores) + c(-1, 1) * sd(scores) / sqrt(length(scores)) ) 
}
bootSpread <- function( scores, conf=.66 ) {
  plugin <- function( d, i ) { mean(d[i]) }

  dist <- boot(data=scores, plugin, R=99) #999 for the publication
  ci <- boot.ci(dist, type = c("bca"), conf=conf)
  return( ci$bca[4:5] ) #The fourth & fifth elements correspond to the lower & upper bound.
}
```

### Individual Components
Here is the basic linear rolling graph.  It doesn't require much specification, and will work with a wide range of approriate datasets.  This first (unpublished) graph displays all components.

```{r Figure2IndividualBasic}
dsLinear <- CountyMonthBirthRate2005Version[CountyMonthBirthRate2005Version$CountyName=="oklahoma", ]

dsLinear <- AugmentYearDataWithMonthResolution(dsLinear=dsLinear, dateName="Date")

portfolioCartesian <- AnnotateData(dsLinear, dvName="BirthRate", centerFunction=median, spreadFunction=hSpread)

CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  changePointLabels = "Bombing Effect"
)
```

The version for the manuscript was tweaked to take advantage of certain features of the dataset.

```{r Figure2IndividualStylized}
lightTheme <- ggplot2::theme(
  axis.title          = element_text(color="gray60", size=9),
  axis.text.x         = element_text(color="gray80", hjust=0),
  axis.text.y         = element_text(color="gray80"),
  axis.ticks.length   = grid::unit(0, "cm"), #g <- g + theme(axis.ticks=element_blank())
  axis.ticks.margin   = grid::unit(.00001, "cm"),
  panel.grid.minor.y  = element_line(color="gray99", size=.1),
  panel.grid.major    = element_line(color="gray95", size=.1),
  panel.margin        = grid::unit(c(0, 0, 0, 0), "cm"),
  plot.margin         = grid::unit(c(0, 0, 0, 0), "cm")
)
dateSequence <- seq.Date(from=as.Date("1990-01-01"), to=as.Date("1999-01-01"), by="years")
xScale       <- scale_x_date(breaks=dateSequence, labels=scales::date_format("%Y"))
xScaleBlank  <- scale_x_date(breaks=dateSequence, labels=NULL) #This keeps things proportional down the three frames.
# yScale <- scale_y_continuous(breaks=5:7)
# yExpand <- expand_limits(y=c(5, 7))

topPanel <- CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate",
  changePointLabels = "Bombing Effect", 
  drawRollingBand = FALSE, 
  drawRollingLine = FALSE
)

middlePanel <- CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate",
  changePointLabels = "", 
  drawRollingBand = FALSE, 
  drawJaggedLine = FALSE
)

bottomPanel <- CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate", 
  changePointLabels = "", 
  drawJaggedLine = FALSE
)

topPanel <- topPanel + xScale + lightTheme #+ yScale + yExpand
middlePanel <- middlePanel + xScale + lightTheme #+ yScale + yExpand
bottomPanel <- bottomPanel + xScaleBlank + lightTheme #+ yScale + yExpand
```

### Combined Panels
This is what it looks like when all three stylized panels are combined.

```{r Figure2Combined, fig.height=4.8} 
#out.height=4.8, out.width=6.5,
vpLayout <- function(x, y) { viewport(layout.pos.row=x, layout.pos.col=y) }
grid.newpage()
pushViewport(viewport(layout=grid.layout(3,1)))
print(topPanel, vp=vpLayout(1, 1))
print(middlePanel, vp=vpLayout(2, 1))
print(bottomPanel, vp=vpLayout(3, 1))
popViewport()
```

## Figure 4: Carteisan Rolling
Re-portrayal of the linear plot of the GFR time series in Oklahoma County.  This will probably be cut from the manuscript.  But here is the basic & tweaked version.

```{r Figure4Basic}
cartesianPeriodicSimple <- CartesianPeriodic(
  portfolioCartesian$dsLinear, 
  portfolioCartesian$dsPeriodic, 
  xName = "Date", 
  yName = "BirthRate",
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  changePointLabels = "Bombing Effect",
  yTitle = "General Fertility Rate",
  drawPeriodicBand = FALSE
)
print(cartesianPeriodicSimple) #Print isn't necessary, but it makes my intention a little clearer.
```

```{r Figure4Stylized}
# yScale <- scale_y_continuous(breaks=5:7)
# yExpand <- expand_limits(y=c(5, 7))

print(cartesianPeriodicSimple + xScale + lightTheme) # + yScale + yExpand
```

## Figure 5: Carteisan Periodic 
Carteisan plot of the GFR time series data in Oklahoma County, with H-spread Bands superimposed.

```{r Figure5Basic}
cartesianPeriodic <- CartesianPeriodic(
  portfolioCartesian$dsLinear, 
  portfolioCartesian$dsPeriodic, 
  xName = "Date", 
  yName = "BirthRate",
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  changePointLabels = "Bombing Effect",
  yTitle = "General Fertility Rate",
  drawPeriodicBand = TRUE #The only difference from the simple linear graph above
)
print(cartesianPeriodic)
```
```{r Figure5Stylized}
cartesianPeriodic <- cartesianPeriodic + xScale + lightTheme # + yScale + yExpand
print(cartesianPeriodic)
```

## Figure 6: Polar Periodic
Wrap Around Time Series (WATS Plot) of the Oklahoma City GFR data, 1990-1999

```{r Figure6, fig.height=6, fig.width=6}
portfolioPolar <- PolarizeCartesian( 
  dsLinear = portfolioCartesian$dsLinear, 
  dsStageCycle = portfolioCartesian$dsStageCycle, 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  plottedPointCountPerCycle = 7200
)

windows.options(antialias = "cleartype")
grid.newpage()
PolarPeriodic(
  dsLinear = portfolioPolar$dsObservedPolar, 
  dsStageCycle = portfolioPolar$dsStageCyclePolar, 
  yName = "Radius",
  stageIDName = "StageID", 
  drawPeriodicBand = FALSE,
  cardinalLabels = c("Jan1", "Apr1", "July1", "Oct1")
)

```
## Figure 7: WATS and Cartesian
Wrap Around Time Series (WATS Plot) of the Oklahoma City GFR data, 1990-1999

```{r Figure7, fig.height=6.5*2/3}
portfolioPolar <- PolarizeCartesian(
  dsLinear = portfolioCartesian$dsLinear, 
  dsStageCycle = portfolioCartesian$dsStageCycle, 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  plottedPointCountPerCycle = 7200
)

grid.newpage()
pushViewport(viewport(
  layout=grid.layout(
    nrow = 2, ncol = 2, respect = TRUE, 
    widths = unit(c(1,1), c("null", "null")), 
    heights = unit(c(1,.5), c("null", "null"))
  ), 
  gp = gpar(cex=1, fill=NA)
))

## Create top left panel
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
topLeftPanel <- PolarPeriodic(  
  dsLinear = portfolioPolar$dsObservedPolar, 
  dsStageCyclePolar = portfolioPolar$dsStageCyclePolar, 
  yName = "Radius", 
  stageIDName = "StageID", #graphCeiling=7, 
  cardinalLabels = c("Jan1", "Apr1", "July1", "Oct1")
)
upViewport()

## Create top right panel
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
topRighttPanel <- PolarPeriodic(
  dsLinear = portfolioPolar$dsObservedPolar, 
  dsStageCyclePolar = portfolioPolar$dsStageCyclePolar, 
  yName = "Radius", 
  stageIDName = "StageID", #graphCeiling=7, 
  drawObservedLine = FALSE,
  cardinalLabels = c("Jan1", "Apr1", "July1", "Oct1"), 
  originLabel = NULL
)
upViewport()

## Create bottom panel
pushViewport(viewport(layout.pos.col=1:2, layout.pos.row=2, gp=gpar(cex=1)))
print(cartesianPeriodic, vp=vpLayout(x=1:2, y=2)) #Print across both columns of the bottom row.
upViewport()
```

## Figure 8: County Comparison
This figure compares Oklahoma County against the (other) largest urban counties.
```{r Figure8, fig.height=6.5}
dsLinearAll <- AugmentYearDataWithMonthResolution(dsLinear=CountyMonthBirthRate2005Version, dateName="Date")

#Identify the average size of the fecund population
plyr::ddply(dsLinearAll, "CountyName", summarize, Mean=mean(FecundPopulation))


GraphCountyComparison <- function( rowLabel="", countyName="oklahoma", spreadFunction=hSpread, changeMonth=as.Date("1996-02-15") ) {
  dsLinear <- CountyMonthBirthRate2005Version[CountyMonthBirthRate2005Version$CountyName==countyName, ]
  dsLinear <- AugmentYearDataWithMonthResolution(dsLinear=dsLinear, dateName="Date")
  portfolioCartesian <- AnnotateData(dsLinear, dvName="BirthRate", centerFunction=median, spreadFunction=spreadFunction)
  portfolioPolar <- PolarizeCartesian(dsLinear=portfolioCartesian$dsLinear, dsStageCycle=portfolioCartesian$dsStageCycle, yName="BirthRate", stageIDName="StageID", plottedPointCountPerCycle=7200)
  cartesianPeriodic <- CartesianPeriodic(portfolioCartesian$dsLinear, portfolioCartesian$dsPeriodic, xName="Date", yName="BirthRate", stageIDName="StageID", changePoints=changeMonth, changePointLabels=""  )
  
  pushViewport(viewport(
    layout=grid.layout(nrow=1, ncol=3, respect=F, widths=unit(c(2,1,3), c("line", "null", "null"))), 
    gp=gpar(cex=1, fill=NA)
  ))
  pushViewport(viewport(layout.pos.col=1))
  grid.rect(gp=gpar(fill="gray90", col=NA))
  grid.text(rowLabel, rot=90)
  popViewport()
  
  pushViewport(viewport(layout.pos.col=2))
  polarPeriodic <- PolarPeriodic(dsLinear=portfolioPolar$dsObservedPolar, dsStageCyclePolar=portfolioPolar$dsStageCyclePolar, drawObservedLine=FALSE, yName="Radius", stageIDName="StageID", originLabel=NULL)
  popViewport()
  
  pushViewport(viewport(layout.pos.col=3))
  print(cartesianPeriodic + xScale + lightTheme, vp=vpLayout(x=1, y=1))
  popViewport()
  popViewport() #Finish the row
}

counties <- c("rogers", "tulsa", "oklahoma", "cleveland", "comanche")

grid.newpage()
pushViewport(viewport(layout=grid.layout(nrow=length(counties), ncol=1), gp=gpar(cex=1, fill=NA)))
for( i in seq_along(counties) ) {
  pushViewport(viewport(layout.pos.col=1, layout.pos.row=i))
  GraphCountyComparison(countyName=counties[i], rowLabel=counties[i])
  popViewport()
}
popViewport()
```


## Figure 9: Error Band Comparison
This figure demonstrates that WATS accommodates many types of error bands.
```{r Figure9, fig.height=6.5 * 4/5}
spreads <- c("hSpread", "fullSpread", "seSpread", "bootSpread")
grid.newpage()
pushViewport(viewport(layout=grid.layout(nrow=length(spreads), ncol=1), gp=gpar(cex=1, fill=NA)))
for( i in seq_along(spreads) ) {
  pushViewport(viewport(layout.pos.col=1, layout.pos.row=i))
  GraphCountyComparison(spreadFunction=get(spreads[i]), rowLabel=spreads[i])
  upViewport()
}
upViewport()
```

## Session Info
For the sake of documentation and reproducibility, the current vignette was build on a system using the following software.

```{r session_info, echo=FALSE}
cat("Report created by", Sys.info()["user"], "at", strftime(Sys.time(), "%c, %z"))
sessionInfo()
```
