<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Assesing OKC Fertility with Intercensal Estimates}
-->

# Assesing OKC Fertility with Intercensal Estimates
The MBR manuscript demonstrates WATS plots with data prepared for Rodgers, St. John, & Coleman (2005).  In that paper and the MBR manuscript, the denominator of the GFR (General Fertility Rate) is the initial 1990 county population of females ages 15-44.  

This vignette uses slightly different Census estimates.  The intercensal population estimates (for females ages 15-44) are used for Jan 1990, Jan 1991, Jan 1992, ..., Jan 2000.  Then linear interpolation is used to complete the remaining 11 months of each year.  These monthly estimates become the denominator of each county's monthly GFR.

```{r set_root_directory, echo=FALSE, results='hide'} 
#It works better if the root directory is set in its own chunk.
library(knitr)
knitr::opts_knit$set(root.dir = "../")
```

```{r set_options, echo=FALSE, results='hide'}
library(base)
library(utils)
library(stats)
library(grDevices)
library(grid)
library(plyr) 
library(scales)
library(ggplot2) 
library(boot) 
library(Wats)

knitr::opts_chunk$set(
    comment=NA, 
    tidy=FALSE,
    fig.width=6.5, 
    fig.height=1.6, 
    out.width="600px", #This affects only the markdown, not the png.  Height will scale appropriately.
    fig.path='figure_okc_fertility_intercensal_rmd/',
    dpi=200
)

base::options(width=120) #So the output is 50% wider than the default.
grDevices::windows.options(antialias = "cleartype")
```

```{r Definitions}
changeMonth <- base::as.Date("1996-02-15") #as.Date("1995-04-19") + lubridate::weeks(39) = "1996-01-17"

vpLayout <- function(x, y) { grid::viewport(layout.pos.row=x, layout.pos.col=y) }

fullSpread <- function( scores ) { 
  return( base::range(scores) ) #A new function isn't necessary.  It's defined in order to be consistent.
}
hSpread <- function( scores ) { 
  return( stats::quantile(x=scores, probs=c(.25, .75)) ) 
}
seSpread <- function( scores ) { 
  return( base::mean(scores) + base::c(-1, 1) * stats::sd(scores) / base::sqrt(base::length(scores)) ) 
}
bootSpread <- function( scores, conf=.66 ) {
  plugin <- function( d, i ) { base::mean(d[i]) }

  dist <- boot::boot(data=scores, plugin, R=99) #999 for the publication
  ci <- boot::boot.ci(dist, type = c("bca"), conf=conf)
  return( ci$bca[4:5] ) #The fourth & fifth elements correspond to the lower & upper bound.
}

darkTheme <- ggplot2::theme(
  axis.title          = ggplot2::element_text(color="gray30", size=9),
  axis.text.x         = ggplot2::element_text(color="gray30", hjust=0),
  axis.text.y         = ggplot2::element_text(color="gray30"),
  axis.ticks.length   = grid::unit(0, "cm"),
  axis.ticks.margin   = grid::unit(.00001, "cm"),
#   panel.grid.minor.y  = element_line(color="gray95", size=.1),
#   panel.grid.major    = element_line(color="gray90", size=.1),
  panel.margin        = grid::unit(c(0, 0, 0, 0), "cm"),
  plot.margin         = grid::unit(c(0, 0, 0, 0), "cm")
)
lightTheme <- darkTheme + ggplot2::theme(
  axis.title          = ggplot2::element_text(color="gray80", size=9),
  axis.text.x         = ggplot2::element_text(color="gray80", hjust=0),
  axis.text.y         = ggplot2::element_text(color="gray80"),
  panel.grid.minor.y  = ggplot2::element_line(color="gray99", size=.1),
  panel.grid.major    = ggplot2::element_line(color="gray95", size=.1)
)
dateSequence <- base::seq.Date(from=base::as.Date("1990-01-01"), to=base::as.Date("1999-01-01"), by="years")
xScale       <- ggplot2::scale_x_date(breaks=dateSequence, labels=scales::date_format("%Y"))
xScaleBlank  <- ggplot2::scale_x_date(breaks=dateSequence, labels=NULL) #This keeps things proportional down the three frames.
```

## Section 1: Cartesian Rolling plot

```{r CartesianRolling, fig.height=4.8}
dsLinear <- utils::read.csv("./Datasets/CountyMonthBirthRate2014Version.csv", stringsAsFactors=FALSE)
dsLinear$Date <- base::as.Date(dsLinear$Date)
dsLinear <- dsLinear[dsLinear$CountyName=="oklahoma", ] 

# Uncomment this line to use the version built into the package.  By default, it uses the
# CSV to promote reproducible research, since the CSV format is more open and accessible to more software.
# dsLinear <- CountyMonthBirthRate2014Version[CountyMonthBirthRate2014Version$CountyName=="oklahoma", ] 

dsLinear <- Wats::AugmentYearDataWithMonthResolution(dsLinear=dsLinear, dateName="Date")

portfolioCartesian <- Wats::AnnotateData(dsLinear, dvName="BirthRate", centerFunction=stats::median, spreadFunction=hSpread)

topPanel <- Wats::CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate",
  changePointLabels = "Bombing Effect", 
  drawRollingBand = FALSE, 
  drawRollingLine = FALSE
)

middlePanel <- CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate",
  changePointLabels = "", 
  drawRollingBand = FALSE, 
  drawJaggedLine = FALSE
)

bottomPanel <- Wats::CartesianRolling(
  dsLinear = portfolioCartesian$dsLinear, 
  xName = "Date", 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  yTitle = "General Fertility Rate", 
  changePointLabels = "", 
  drawJaggedLine = FALSE
)

topPanel <- topPanel + xScale + darkTheme
middlePanel <- middlePanel + xScale + darkTheme
bottomPanel <- bottomPanel + xScaleBlank + darkTheme

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout=grid::grid.layout(3,1)))
print(topPanel, vp=vpLayout(1, 1))
print(middlePanel, vp=vpLayout(2, 1))
print(bottomPanel, vp=vpLayout(3, 1))
grid::popViewport()
```

## Section 2: Carteisan Periodic 
Carteisan plot of the GFR time series data in Oklahoma County, with H-spread Bands superimposed.

```{r CartesianPeriodic}
cartesianPeriodic <- Wats::CartesianPeriodic(
  portfolioCartesian$dsLinear, 
  portfolioCartesian$dsPeriodic, 
  xName = "Date", 
  yName = "BirthRate",
  stageIDName = "StageID", 
  changePoints = changeMonth, 
  changePointLabels = "Bombing Effect",
  yTitle = "General Fertility Rate",
  drawPeriodicBand = TRUE #The only difference from the simple linear graph above
)
cartesianPeriodic <- cartesianPeriodic + xScale + darkTheme 
print(cartesianPeriodic)
```

## Section 3: Polar Periodic
Wrap Around Time Series (WATS Plot) of the Oklahoma City GFR data, 1990-1999

```{r PolarPeriodic, fig.height=6.5*2/3}
portfolioPolar <- Wats::PolarizeCartesian(
  dsLinear = portfolioCartesian$dsLinear, 
  dsStageCycle = portfolioCartesian$dsStageCycle, 
  yName = "BirthRate", 
  stageIDName = "StageID", 
  plottedPointCountPerCycle = 7200
)

grid::grid.newpage()
grid::pushViewport(grid::viewport(
  layout=grid::grid.layout(
    nrow = 2, ncol = 2, respect = TRUE, 
    widths = grid::unit(c(1,1), c("null", "null")), 
    heights = grid::unit(c(1,.5), c("null", "null"))
  ), 
  gp = grid::gpar(cex=1, fill=NA)
))

## Create top left panel
grid::pushViewport(grid::viewport(layout.pos.col=1, layout.pos.row=1))
topLeftPanel <- Wats::PolarPeriodic(  
  dsLinear = portfolioPolar$dsObservedPolar, 
  dsStageCyclePolar = portfolioPolar$dsStageCyclePolar, 
  yName = "Radius", 
  stageIDName = "StageID", #graphCeiling=7, 
  cardinalLabels = c("Jan1", "Apr1", "July1", "Oct1")
)
grid::upViewport()

## Create top right panel
grid::pushViewport(grid::viewport(layout.pos.col=2, layout.pos.row=1))
topRighttPanel <- Wats::PolarPeriodic(
  dsLinear = portfolioPolar$dsObservedPolar, 
  dsStageCyclePolar = portfolioPolar$dsStageCyclePolar, 
  yName = "Radius", 
  stageIDName = "StageID", #graphCeiling=7, 
  drawObservedLine = FALSE,
  cardinalLabels = c("Jan1", "Apr1", "July1", "Oct1"), 
  originLabel = NULL
)
grid::upViewport()

## Create bottom panel
grid::pushViewport(grid::viewport(layout.pos.col=1:2, layout.pos.row=2, gp=grid::gpar(cex=1)))
print(cartesianPeriodic, vp=vpLayout(x=1:2, y=2)) #Print across both columns of the bottom row.
upViewport()
```

## Section 4: Confirmatory Analysis of Interrupted Time Series
This section uses an approach advocated by McLeod, Yu, & Mahdi (2011), which is consistent other articles, including Rodgers et al. (2005).  There are two trends that are de-seasonalized.  The first is the 'classic' approach which uses the observed trend line.  The second is a smoothed version, where a loess is passed through the observed data; this smoothed line is then de-seasonalized.  Both approach give comparable conclusions.  The post-bombing fertility is significantly higher  than the pre-bombing fertility (ie, the `step` coefficient is significantly more positive).

```{r ConfirmatoryFrequentist, fig.height=6.5}
dsLinear <- Wats::AugmentYearDataWithMonthResolution(dsLinear=dsLinear, dateName="Date")

tsData <- stats::ts(
  data = dsLinear$BirthRate, 
  start = as.integer(dsLinear[1, c("Year", "Month")]), 
  end = as.integer(dsLinear[nrow(dsLinear), c("Year", "Month")]),
  frequency = 12L
)

#Create unsmoothed and smoothed version
seasonalClassic <- stats::decompose(tsData)
plot(seasonalClassic)

#Watch out, the 2nd & 3rd columns are swapped, compared to `decompose()`
seasonalLoess <- stats::stl(x = tsData, s.window = "periodic") 
plot(seasonalLoess)


# Seasonality is accounted for without a smoother
lag <- 1 #Significant for many different values of lag, including 1
y <- seasonalClassic$trend[(lag+1):length(seasonalClassic$trend)]
y1 <- seasonalClassic$trend[1:(length(seasonalClassic$trend)-lag)]
step <- c(rep(0, times=sum(dsLinear$StageID==1)-lag), rep(1, times=sum(dsLinear$StageID==2)))
dsClassic <- data.frame(y=y, y1=y1, step=step)
rm(lag, y, y1, step)
fitClassic <-  glm(y ~ 1 + step + y1, data=dsClassic)
summary(fitClassic)

#Seasonality is accounted for after a loess is fit through it.
lag <- 1 #Significant for many different values of lag, including 1
trendLineLoess <- as.numeric(seasonalLoess$time.series[, 2])
y <- trendLineLoess[(lag+1):length(trendLineLoess)]
y1 <- trendLineLoess[1:(length(trendLineLoess) - lag)]
step <- c(rep(0, times=sum(dsLinear$StageID==1)-lag), rep(1, times=sum(dsLinear$StageID==2)))
dsLoess <- data.frame(y=y, y1=y1, step=step)
rm(lag, y, y1, step)
fitLoess <-  glm(y ~ 1 + step + y1, data=dsLoess)
summary(fitLoess)
```

## Section 5: Confirmatory Analysis of Interrupted Time Series with Bayes Factors
A similar set of Bayesian analyses support the claim that post-bombing fertility is higher.  The `BayesSingleSub` package by De Vries and Morey.  The third number in each row describes the test for a mean different.  The package also has tests that indicate the post-bombing slope may be more positive than the pre-bombing slope.  Significance is sugested with a Bayes Factors is less than one (and therefore its log is negative).

```{r ConfirmatoryBayesFactors, fig.height=6.5}
# Seasonality is accounted for without a smoother
beforeClassic <- seasonalClassic$trend[dsLinear$StageID==1]
afterClassic <- seasonalClassic$trend[dsLinear$StageID==2]
mcmcRepCount <- 1000#000
(gClassic <- BayesSingleSub::trendtest.Gibbs.AR(beforeClassic[!is.na(beforeClassic)], afterClassic[!is.na(afterClassic)], return.chains=F, iterations=mcmcRepCount, progress=interactive()))
(mcClassic <- BayesSingleSub::trendtest.MC.AR(beforeClassic[!is.na(beforeClassic)], afterClassic[!is.na(afterClassic)], iterations=mcmcRepCount, progress=interactive()))
summary(mcClassic)
# coda::gelman.diag(g$chains) #it needs multiple chains

#Seasonality is accounted for after a loess is fit through it.
beforeLoess <- seasonalLoess$time.series[dsLinear$StageID==1, 2]
afterLoess <- seasonalLoess$time.series[dsLinear$StageID==2, 2]
BayesSingleSub::trendtest.Gibbs.AR(beforeLoess, afterLoess, iterations=mcmcRepCount, progress=interactive())
BayesSingleSub::trendtest.MC.AR(beforeLoess, afterLoess, iterations=mcmcRepCount, progress=interactive())

```

## Session Info
For the sake of documentation and reproducibility, the current vignette was build on a system using the following software.

```{r session_info, echo=FALSE}
base::cat("Report created by", base::Sys.info()["user"], "at", base::strftime(base::Sys.time(), "%c, %z"))
utils::sessionInfo()
```

## References
* McLeod, A.I., Yu, H., & Mahdi, E. (2011). [Time series analysis with R.](http://www.stats.uwo.ca/faculty/aim/tsar/tsar.pdf). *Handbook of Statistics*, Volume 30, Elsevier. 
* Rodgers, J. L., St. John, C. A. & Coleman R.  (2005).  [Did Fertility Go Up after the Oklahoma City Bombing?  An Analysis of Births in Metropolitan Counties in Oklahoma, 1990-1999](http://www.ncbi.nlm.nih.gov/pubmed/16463916).  *Demography, 42*, 675-692.
*De Vries, R. M. & Morey, R. D. (submitted). Bayesian hypothesis testing Single-Subject Data.
Psychological Methods.
